---
title: "Time series - decomposition"
output: html_notebook
---

data should be extracted
12 interval monthly data about number of passngers for 12 years 
```{r}
data("AirPassengers")
AP <- AirPassengers
str(AP)
head(AP)
ts(AP, frequency = 12, start = c(1949,1))
attributes(AP)
plot(AP)
```

there is an increasing trend but also some seasonality can be seen. 
the spans in initial series is smaller than toward which make time series non stationary 
therefore we apply log transformation

```{r}
AP <- log(AP)
plot(AP)
```

The fluctuation now is less between spans


Decomposition
a summary of time series
figure is for estimated the seasonal figure only
```{r}
decomp <- decompose(AP)
decomp$figure
plot(decomp$figure,
     type='b',
     xlab='Month',
     ylab='seasonality Index',
     col='blue',
     las=2)
```
as it is shown the June until september has the maximum avg of passeger by 20% more than total avg


for showing trend, seasonal, and random simply
```{r}
plot(decomp)
```

package 
```{r}
install.packages("forecast")
```



Autoregression integerated moving average 
ACF and PACF plots 
ljung-box test


```{r}
library(forecast)
model <- auto.arima(AP)
model
attributes(model)
model$coef
```
(0,1,1) p:ar order, d:degree of differencing, q:Moving average order

ACF and PACF plots
```{r}
acf(model$residuals, main='Correlogram')
pacf(model$residuals, main='partial Correlogram')
```
autocorrelation bars havenot excessed the significant bars. only one varible has touched the significant bar. 
in partial correlationgram all has been good. for testing that unique value we use below test. 


Ljung-box test 
```{r}
Box.test(model$residuals, lag=20, type = 'Ljung-Box')
```
since the p value is more than 0.05 there is little chance for nonzero autocorrelation or we dont have any autocorrelation 

just for conefirming no problem with the model 
residual plot 
we are looking a bell shape histogram which is not based on frequency and is beased on density 
```{r}
hist(model$residuals,
     col='red',
     xlab = 'Error',
     main='Histogram of residuals',
     freq = FALSE)
lines(density(model$residuals))
```
since there is no problem with model we can forecast 

Forecast
forecast for next 48 month 
```{r}
library(ggplot2)
f <- forecast(model,48)
autoplot(f)
accuracy(f)
```
dark 80% confidence and light is 95% confidence 
standard accuracy measures have been computed 













